Social Visual Analytics: DevOps for Data Scientists

### 2015-03-20 conclusion.tex
working

We implemented these ideas in the RCloud prototype.
Experience with the prototype provides evidence that data
science teams and the organizations in which they work
benefit from having such capabilities.

This work prototype a step toward practical ``DevOps for data
science'' and offers reproducible experiments.
Possible next steps are to support cross-language development,
to incorporate richer recommendation techniques, to provide fine-grained
information security, and to improve the usability of the human interface.

>>>>

Building from the extensive previous literature identifying
requirements for future visual analytics systems, our goal was
to gain experience and further insight by implementing and
deploying a prototype with a community of working data scientists.
Experience with the RCloud prototype provides evidence that data
science teams and the organizations in which they work
benefit from having such capabilities that support a
broad view of the visual analytics process.

Rcloud proposes a step toward practical ``DevOps for data
science'' with reproducible experiments.
Possible next steps are to support cross-language development,
to incorporate richer recommendation techniques, to provide fine-grained
information security, and to improve the usability of the human interface.

### 2015-03-20 discussion.tex
working

\section{Discussion, Lessons and Limitations}
\label{sec:limitations}

@@

\carlos{All this stuff needs to appear in the paper much earlier.}

Experience with deploying RCloud in a community of data analysts
(``hackers'') has provided some insight about whether the proposed
requirements were appropriate, and how well they were met.
\stephen{The term {\it hackers} may be too tainted for us to use.}
% I know the following is true but also feel we need to let the work speak for itself. SCN
%This experience may be informative to researchers and implementers
%of future systems supporting collaborative visual analytics.

Returning to Heer and Agrawala's design considerations, our experience
underscores the relevance of their proposals, and indicates areas
for further exploration. We briefly examine these aspects of the
visual analytics environment.

{\bf Shared artifacts and artifact histories} are a central feature
of RCloud. The prototype aimed at making it convenient to collaborate
on experimental workbooks and the resulting analyses. We observed
that hackers readily shared ideas through RCloud. They used RCloud
to discuss data analysis, programming problems, and to publish results
by sharing URLs instead of instead of embedding screenshots in
PowerPoint. Artifact histories are provided through git. It is fair
to criticize that it is not human-friendly, but we assume RCloud
hackers are already familiar with it.

{\bf Modularity and granularity.} The ability to segment work into
independent units (modularity) is a key to working in teams. It is
further an advantage if the units may be kept small, and team members
can realize benefits that are at least proportional to their work
on these units (granularity). RCloud's notebook and versioning
capabilities allow work to be divided into units as fine as
the underlying language allows, and versioning encourages making
incremental changes at low cost and without disrupting the work
of others.

{\bf View sharing, bookmarking.} Most resources in RCloud are named
and accessed as URLs. This proved to be an effective mechanism for
sharing, and for analyses to be integrated with external processes
and systems.

{\bf Discussion.} Annotation and commenting was another central goal.
Commenting is supported through github, but our hackers found it
awkward and did not exercise it as much as we expected.
making it a bit awkward and
not as widely used as we hoped.  More elaborate annotation was not
implemented, and remains as future work.
An interesting question is, to what extent should application users
be able to make annotations and operate similar tools in published
notebooks without coding and being exposed to the hackers's view?
Our prototype has not addressed this problem yet.

{\bf Content export} is not a capability we worked to support,
but R already has many packages for this. Our goal was to devise
ways of working where explicit export is less necessary.
However it is desirable to integrate results
through web protocols such as JSON. We did assume that RCloud
notebooks need to play well in the ecosystem of other tools,
and this worked out as intended, though we learned it also creates
a temptation to absorb features almost without limit. (Should
RCloud support templates for personal web pages and graphical
editing? Should it incorporate an email client and emacs?
Should it be able to push results to Facebook?)

{\bf Social-psychological incentives} and {\bf voting and ranking}
are supported through starring. This mechanism was employed often
by our test audience. An obvious next step would be to enhance
recommendations via relationships discovered by static and
dynamic code analysis. This should be done both within and across
collections of scripts (the latter being similar to Vistrails'
enhanced recommendations by clustering multiple workflows).
It should be valuable to hackers to know which packages are
frequently used together, or appear with a given record type
or data feed. In general, though, trivial or passive mechanisms
to collect data for recommendations are essential.

{\bf Group management, size and diversity.} This area lacks sufficient
support in RCloud, but is clearly important to working in teams.
We rely on existing administrative and social processes and conventions,
outside RCloud, to manage accounts and groups. This capability
could likely benefit from integrating social media to track
identities and groups and to maintain communication channels.

{\bf Curation.} Even without formal group management, users are
able to curate groups of related notebooks using the folder
system. This is particularly effective for collecting and
distributing training materials.

\subsection{Limitations}

Because we developed these ideas while creating a prototype,
we did not foresee some of the requirements that emerged after
people started working with it.

{\bf Multiple language support.}
We did not anticipate having to back off so soon on the assumption that
programming is done mainly in R.  In retrospect, it is clear that many
data science problems are solved with cross-language approaches
(for example, scikit-learn in Python, weka in Java).

{\bf Versioning.}
Some important aspects of the environment are more difficult to manage
in RCloud than in conventional systems.  RCloud does not explicitly
separate the development and deployment environments. Actually, every
version of every prototype is shared by default. Although this encourages
collaboration on work in progress, it also quickly exposes errors such as
misconfiguration, mismatches between versions of packages, and programming
errors that can affect production websites unintentionally.
This cannot be avoided entirely, but the power of sharing is often worth
this price (as proven true with distributed systems in general).
Still, effective control over versioning when sharing 
problem needs more attention if the environment is
going to be scaled up to a wider number of users with fewer
informal channels to coordinate work.
Versioning must be managed in several places: in scripts, in the R environment
such as the installed R libraries and packages, and in the external
environment such as the operating system and even protocols spoken by
remote services. At one end of the scale, we have full control over the
versioning of scripts via git, along with conventions to name stable or
working versions of scripts or versions known to work together.
The R environment is under the control of its package manager.
It is at least possible for RCloud to access information about
package versions and configurations, but support for checking compatibility
and maintaining multiple versions in the same environment is not strong.
At the other end of the scale, there is not much reason to expect version
control for the external environment. Most programmers rely on clean living
and a conservative approach to upgrades.

On top of this, RCloud allows hackers to fork experiments to try
new ideas quickly. So far we have not done much more to address the
problem of having a lot of bits and pieces of code and data lying
around, though we have provided a structure in which it should be
easier to find them and to apply algorithms and metadata to organize them.
\stephen{Why should the reader believe that?}

@@

{\bf Caching.}
An important consideration is how and where to introduce caching
in RCloud's distributed computation model. Caching can dramatically
improve performance in a way that is otherwise transparent to
application program semantics \cite{Callahan:2006:VVM, Guo:2010:TPI}.
Though caching is usually implicit, in some environments, such as
Vistrails, programmers may also have some explicit control over cache
management. This may be desirable to ensure the repeatability of
computations that rely on volatile or unreliable data sources.
For example, the stock ticker use case is one for which the data
might temporarily become unavailable, so caching could improve
reliability and consistency of results.  Alternatively, in situations
involving relatively expensive analysis (for example, analysis of large
text corpora, clustering of multivariate time series, deep learning, etc.)
caching may be essential to provide adequate performance. Currently we would
program this in RCloud by explicitly saving the analysis in a persistent database,
but it would be desirable to implement this in a general purpose associative
cache, not in application-dependent libraries.  We envision cache management
being implemented in a new middle layer not in RCloud's architecture today.

>>>>

\section{Discussion, Lessons and Limitations}
\label{sec:discussion}

@@

Experience with deploying RCloud in a community of data analysts
(``hackers'') gave us some insight into whether the proposed
requirements are appropriate, and how well they were met.

Our experience underscored the relevance of Heer and Agrawala's design
considerations, and indicated areas for further exploration. We adopt
their taxonomy for the following discussion.

{\bf Shared artifacts and artifact histories} are a central feature
of RCloud, through its shared workbooks of experiments and analyses.
We observed that hackers readily shared work through RCloud. They
used it to discuss ideas and results in data analysis, to ask for help
with programming problems, and to publish results by sharing URLs.
Artifact histories are provided through a version management system (git).
\stephen{Do analysts type raw git commands?}

{\bf Modularity and granularity.} The ability to partition work into
independent units (modularity) is a key to working productively in teams.
It is good if the units can be kept small, so team members can realize
benefits at least proportional to the work on the units (granularity).
RCloud's notebook and versioning capabilities allow work to be divided
into units as fine as the underlying language allows, and versioning
encourages making incremental changes at low cost and without disrupting
the work of others.

{\bf View sharing, bookmarking.} Most resources in RCloud are named
and accessed as URLs. This proved to be an effective mechanism for
sharing and for integrating analyses with external processes and systems.
It is particularly advantageous for work to be shared as URLs that
provide access to code and experiments, instead of by pasting static
screenshots into documents and presentations.

{\bf Discussion.} Annotation and commenting was another central goal.
Commenting is supported through github, but our hackers found it
awkward and did not exercise it as much as we expected.
An interesting question is, to what extent should application users
be able to make annotations in published notebooks without coding
and being exposed to the hackers's view? The design of
more elaborate, integrated annotation remains as future work.

{\bf Content export} is not a capability we aimed at supporting,
and R already has many packages for this. Instead, our goal was to
encourage ways of working without needing as much explicit export.
We did assume that RCloud notebooks should play well in the ecosystem
of other tools, and this worked as intended, though we learned it also
creates a temptation to absorb features almost without limit. (Should
RCloud provide templates for personal web pages and graphical
editing? Should it incorporate an email client and emacs?
Should it push results to Facebook?) Experience shows that interfaces
to external systems and other features almost always win over architectural
purity in the real world.

{\bf Social-psychological incentives} and {\bf voting and ranking}
are supported through starring. This mechanism was employed often
by our test audience. An obvious next step would be to enhance
recommendations using relationships discovered by static and
dynamic code analysis. This may be considered both within and across
collections of scripts (the latter being similar to Vistrails'
enhanced recommendations by clustering multiple workflows).
It seems valuable to know which packages are frequently used together,
or appear in proximity to a certain record types or data feeds.
In general, trivial or passive mechanisms to collect data for
recommendations are essential.

{\bf Group management, size and diversity.} This area needs better
support in RCloud, but is clearly important to working in teams.
We rely on external administrative processes and social conventions
to manage accounts and groups. Rcloud could benefit greatly from
integrating social media to track identities and groups and to maintain
communication channels, instead of having its own isolated solution.

{\bf Curation.} Even without formal group management, users may
curate groups of related notebooks using file system folders.
We found this particularly effective in collecting and distributing
training materials.

\subsection{Limitations}
Because we developed our ideas while simultaneously creating a prototype,
we did not foresee some of the requirements that emerged after
people started working with RCloud.

{\bf Multiple language support.}
We did not forsee the demand to relax the assumption that programming
is done mainly in R. In retrospect, it is clear that many data science
problems are solved with cross-language approaches (for example, scikit-learn
in Python, weka in Java).\stephen{Isn't this fixed in some way now?}

{\bf Versioning.}
Some important aspects of the environment are more difficult to manage
in RCloud than in conventional systems. RCloud does not explicitly
separate the development and deployment environments. More than that,
ever version of every prototype is shared by default. Although this encourages
collaboration on work in progress it also quickly exposes errors such as
misconfiguration, mismatches between versions of packages, and programming
errors that can unintentionally affect production websites. Such errors
are difficult to completely avoid, but the power of convenient sharing is often
worth it (as proven true with distributed systems in general).
Still, effective control over versioning when sharing problem needs more
attention if Rcloud would be scaled up to a wider number of users lacking
informal channels to coordinate work.
Versioning needs to be managed in several places: in scripts;
in the R environment such as the installed libraries and packages;
and in the external environment such as the operating system and
protocols spoken by remote services. At one end of the scale, we have
full control over the versioning of scripts via git, along with conventions
to name stable or working versions of scripts, and versions known to work together.
The R environment itself is under the control of its package manager
that has rules to ensure reasonable consistency.
It is at least possible for RCloud to access information about
package versions and configurations, but support for checking compatibility
and maintaining multiple versions in the same environment is not strong.
At the other end of the scale, there is not much reason to expect version
control for the external environment. Most programmers rely on clean living
and a careful approach to system upgrades.

On top of this, RCloud allows or even encourages hackers to fork experiments
to try new ideas quickly. So far we have not done much more to address the
problem of having a lot of bits and pieces of code and data lying around,
though we have provided a framework in which it should be easier to find them
and to apply algorithms and metadata to organize them.
\stephen{I think we need to defend that a bit more.}

@@

{\bf Caching.}
An important consideration is how and where to introduce caching
in RCloud's distributed computation model. Caching can dramatically
improve performance in a way that is otherwise transparent to
application program semantics \cite{Callahan:2006:VVM, Guo:2010:TPI}.
Though caching is usually implicit, in some environments, such as
Vistrails, programmers may also have some explicit control over cache
management. This may be desirable to ensure the repeatability of
computations that rely on volatile or unreliable data sources.
For example, the stock ticker use case is one for which the data
might temporarily become unavailable, so caching could improve
reliability and consistency of results. Alternatively, in situations
involving relatively expensive computation (for example, analysis of large
text corpora, clustering multivariate time series or deep learning algorithms)
caching may be essential to adequate performance. Currently we would
program this in RCloud by explicitly saving an analysis in a persistent database,
but it seems better to implement this capability in a general purpose associative
cache, instead of application-specific libraries. We envision cache management
being implemented in a new middle layer to be added in the future.
